---
---
title: 'Sparse interactions: Simulations'
output:
  pdf_document: 
      highlight: haddock
      toc: true
      toc_depth: 3
  html_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
```

Code used to generate simulated communities for the manuscript "Disentangling key species interactions in diverse and heterogeneous communities: A Bayesian sparse modeling approach." This code generates many simulation runs and is used in figure 2 and supplemental figure S1. 

The companion code BH_Simulations_R1, which generates a single simulation run, has more detailed comments on how each variable is calculated and used.

## Functions

Single Beverton-Holt run

```{r function-BH, eval=FALSE}
BH.run <- function(num.species, N.0, lambda, alpha){
  N.1 <- numeric(length = num.species) # blank vector to be filled in with the for loop
  for(i in 1:num.species){
    comp.sum <- sum(alpha[,i]*N.0)
    N.1[i] <- N.0[i]*lambda[i] / (1 + comp.sum)
  }
  
  return(N.1)
}
```


Multiple runs through time on a given set of species parameters 

```{r function-BH-env, eval=FALSE}
BH.env <- function(df.in, n.runs, t.steps, p.start, env, dem.het = FALSE, lambda.opt = FALSE)
{
  # setting up empty data frame to store the output
  n.species <- length(df.in$species)
  species <- rep(factor(1:n.species), times = n.runs*(t.steps+1))
  run <- rep(1:n.runs, each = n.species*(t.steps+1))
  time <- rep(rep(0:t.steps, each = n.species), times = n.runs)
  pop <- rep(-100, times = n.species*(t.steps+1)*n.runs) 
  run.env <- rep(env, each = n.species*(t.steps+1)) # env conditions in each run
  
  df.out <- data.frame(species, run, run.env, time, pop)
  
  
  # running across each run (environmental condition varies across runs)
  for(r in 1:n.runs){
    # initial populations
    start.rows <- df.out %>% 
      with(run == r & time == 0) %>%
      which()
    start.pop <- with(p.start, pop.size[run.count == r])
    df.out$pop[start.rows] <- start.pop
    
    # calculating lambda for each species in the given environment
    if(lambda.opt == TRUE){ 
      # lambda with environmental optimum
      env.diff <- df.in$z.env - env[r] # diff bw species opt env and current env
      env.width <- 2*df.in$sigma.env # denom for below
      lambda <- df.in$lambda.max * exp(-(env.diff/env.width)^2) 
    }
    else {
      # monotonic lambda-environment interaction
      lambda <- exp(df.in$lambda.mean + df.in$lambda.env*env[r])
    }
    
    # calculating alpha for each species pair in the given environment
    alpha.means <- df.in %>% 
      select(starts_with('alpha') & !starts_with('alpha.env')) # mean alphas
    alpha.env.int <- df.in$alpha.env * env[r] # alpha x env interaction
    alpha <- alpha.means # make a same-sized df
    
    for(i in 1:n.species){
      alpha[i,] <- exp(alpha.means[i,] + alpha.env.int[i])
    }
    
    # simulating the community through time
    for(t in 0:t.steps){
      # where to store the data for the runs
      end.rows <- df.out %>% 
        with(run == r & time == t+1) %>%
        which()
      
      end.pop <- BH.run(n.species, start.pop, lambda, alpha)
      end.pop[end.pop < 0] <- 0 # set negative numbers to zero
      
      # option to add in demographic stochasticity
      if(dem.het == TRUE) {
        end.pop <- rpois(length(end.pop), end.pop)
      }
      
      df.out$pop[end.rows] <- end.pop
      
      start.pop <- end.pop # updating populations for the next time step
    }
  }
  return(df.out)
}
```

Single simulation run. Uses the functions above to simulate plot communities deterministically to near-equilibrium values, perturb species populations, and then look at stochastic population change from that perturbed state to the following time step. Has default inputs for standard function runs.
```{r run-simulation}
run.simulation <- function(x, num.species = 15, num.runs = 500, pre.time.steps = 20, env.variation = 1,
                           pop.start.mean = 80, pop.start.sd = 50, thinned = TRUE, num.runs.thinned = 200,
                           lambda.optimum = FALSE, env.response = 0.5, a.intra = -3.5, a.mean = -7,
                           a.range = 3.5, a.env.sd = 1){
     ### Setting up parameters
     # Setting the environmental condition in each plot
     env.condition <- rnorm(n = num.runs, mean = 0, sd = env.variation)
     
     # Setting lambda values for each species
     if(lambda.optimum == TRUE){
          # optimum environmental condition for species
          # max intrinsic growth
          lambda.max <- runif(n = num.species, min = 1, max = 5) 
          # environmental response
          z.env <- rnorm(n = num.species, mean = 0, sd = env.variation)
          sigma.env <- rexp(n = num.species, rate = 0.5)
     } else { 
          # montonic environmental relationships for species
          lambda.mean <- runif(n = num.species, min = 0, max = 1.5) 
          # environmental response
          lambda.env <- rnorm(n = num.species, mean = 0, sd = env.response)
     }
     
     
     # Setting up alpha values for each species pair 
     
     # setting up competition matrix
     alphas <- runif(num.species, min = a.mean - a.range/2, 
                     max = a.mean + a.range/2)
     alpha <- matrix(rep(alphas, times = num.species), nrow = num.species, ncol = num.species) 
     
     # filling in alphas for intraspecific variaiton
     diag(alpha) <- a.intra
     
     # environmentally-variable alpha terms
     alpha.env <- rnorm(num.species, mean = 0, sd = a.env.sd)
     
     # dataframe of coefficients
     if(lambda.optimum == TRUE){
          df.coef <- data.frame(species = factor(1:num.species), 
                                lambda.max, z.env, sigma.env, 
                                alpha, alpha.env) %>%
               rename_with(.,~gsub("X","alpha.", .x, fixed = TRUE))
     } else { 
          df.coef <- data.frame(species = factor(1:num.species), 
                                lambda.mean, lambda.env, 
                                alpha, alpha.env) %>%
               rename_with(.,~gsub("X","alpha.", .x, fixed = TRUE))
     }
     
     ### Running the functions
     # Setting up the initial populations
     pop.size <- as.integer(rnorm(num.species*num.runs, 
                                  mean = pop.start.mean, 
                                  sd = pop.start.sd))
     pop.size[pop.size < 0] <- 0 # set negative numbers to zero
     run.count <- rep(1:num.runs, each = num.species)
     pop.start <- data.frame(run.count, pop.size)
     
     # Warm-up steps: deterministic simulations to near-equilibrium
     df.eq <- BH.env(df.in = df.coef, 
                     n.runs = num.runs, t.steps = pre.time.steps, 
                     p.start = pop.start, 
                     env = env.condition,
                     dem.het = FALSE,
                     lambda.opt = lambda.optimum)
     
     # Perturbing equilibrium populations with noise.
     pop.eq <- filter(df.eq, time == max(time)) # final time step
     
     pop.eq.dist <- rnorm(length(pop.eq$pop), 
                          mean = pop.eq$pop + 1, 
                          sd = pop.eq$pop + 1) %>% round() # perturbation
     
     pop.eq.dist[pop.eq.dist < 0] <- 0 # no negative population sizes
     pop.eq.dist.df <- data.frame(run.count, pop.size = pop.eq.dist)
     
     # Running the simulation stochastically for one time step from those perturbed population values. 
     # This is the final data used for the sparse model.
     df.result <- BH.env(df.in = df.coef, 
                         n.runs = num.runs, t.steps = 1, 
                         p.start = pop.eq.dist.df, 
                         env = env.condition,
                         dem.het = TRUE,
                         lambda.opt = lambda.optimum)
     
     #If using a thinned treatment with focal individuals, running those simulations and adding them to the data frame. 
     if(thinned == TRUE){
          # choose the focal species from the species that are present in at least 2/3 of the plots
          pop.present <- pop.eq %>% 
               group_by(species) %>%
               dplyr::summarise(
                    n.present = length(which(pop > 1)))
          focal.species.options <- which(pop.present$n.present > (num.runs*.67))
          focal.species <- sample(focal.species.options, size = 1)
          
          df.coef$focal <- 0
          df.coef$focal[focal.species] <- 1
          
          num.focal <- length(focal.species)
          total.runs <- num.runs.thinned*num.focal
          
          species.id <- rep(1:num.species, time = total.runs)
          pop.size <- numeric(num.species*total.runs)
          run.count <- rep(1:(total.runs), each = num.species)
          pop.start.thinned <- data.frame(run.count, pop.size, species.id)
          
          for(i in 1:num.focal){
               runs.start <- (i-1)*num.runs.thinned + 1
               runs.end <- i*num.runs.thinned
               pop.start.thinned$pop.size[run.count %in% runs.start:runs.end & 
                                               species.id == focal.species[i]] <- 1
          }
          
          env.condition.thinned <- rnorm(n = total.runs, 
                                         mean = 0, sd = env.variation)
          
          df.result.thinned <- BH.env(df.in = df.coef, 
                                      n.runs = total.runs, 
                                      t.steps = 1, 
                                      p.start = pop.start.thinned, 
                                      env = env.condition.thinned,
                                      dem.het = TRUE,
                                      lambda.opt = lambda.optimum)
          
          df.result$thinned <- 0
          df.result.thinned$thinned <- 1
          df.result <- df.result %>% rbind(df.result.thinned)
     }
     
     output <- list('parameters' = df.coef, 'simulation' = df.result)
     return(output)
}

```

## Running simulations and storing output
```{r many-simulations, warning=FALSE, message=FALSE}
n.sims <- 10

simulations <- 1:n.sims %>% map(run.simulation)

save(simulations, file=('test_multiple_simulations.rdata'))

```


### Checking that saved output
```{r env-storing, eval=FALSE}
#load('test_multiple_simulations.rdata')
View(simulations)

```


